---
title: "Educational Outcomes of Children in Care in Ireland, 2018-2025"
subtitle: "An Analysis of Educational Attainment, Placement Effects, and Success Factors"
author: "Your Name Here"
date: "`r Sys.Date()`"
format: 
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    geometry: margin=1in
    fontsize: 11pt
bibliography: references.bib
---

```{r setup, include=FALSE}
# Load all required packages
library(tidyverse)
library(purrr)
library(knitr)

# Set knitr options
knitr::opts_chunk$set(
  echo = TRUE,           # Show code
  warning = FALSE,       # Hide warnings
  message = FALSE,       # Hide messages
  fig.width = 8,         # Figure width
  fig.height = 6,        # Figure height
  fig.align = "center"   # Center figures
)

# Source all function scripts
source("scripts/01_download_data.R")
source("scripts/02_data_cleaning.R")
source("scripts/03_analysis_functions.R")
source("scripts/04_visualization_functions.R")
```

\newpage

# Introduction

<!-- 
YOUR CONTENT HERE: 
- Introduce the topic of children in care in Ireland
- Explain why this matters (social significance)
- Brief overview of Tusla and the care system
- State your research questions
- Cite the data source properly
-->

## Background

Children in care in Ireland are among the most vulnerable in society...

## Research Questions

This analysis addresses the following key questions:

1. How do educational outcomes for children in care compare to the general population?
2. What factors predict successful educational and employment outcomes?
3. Does placement type (foster care vs. residential care) affect outcomes?
4. How have outcomes changed from 2018 to 2025?

## Data Source

This analysis uses data from the Central Statistics Office (CSO) publication "Educational Attendance, Attainment and Other Outcomes of Children in Care, 2018-2025" [@cso_children_care_2024]. The dataset contains information on 5,257 children in care in January 2024 and 3,178 children who left care since April 2018.

Data available at: [CSO Website](https://www.cso.ie/en/statistics/education/educationalattendanceandattainmentofchildrenincare/)

\newpage

# Part 1: Data Analysis {#sec-part1}

```{r load-data}
# YOUR CODE HERE: Load your cleaned data
# data <- read_csv("data/processed/cleaned_data.csv")
```

## Data Overview

```{r data-overview}
# YOUR CODE HERE: Show basic data structure
# - Number of observations
# - Number of variables  
# - Variable types
# - Sample of the data
```

<!-- COMMENT: Explain what the dataset contains -->

## Demographic Profile

### Age and Gender Distribution

```{r demographics}
# YOUR CODE HERE: Create demographic summary
# Use your summarize_demographics() function
```

```{r plot-demographics}
# YOUR CODE HERE: Create demographic plots
# Use your plotting functions
```

<!-- COMMENT: Explain what you see in the demographics -->

### Care Characteristics

```{r care-characteristics}
# YOUR CODE HERE: Summarize care types and placements
```

<!-- COMMENT: Explain the care system characteristics -->

## Educational Outcomes Analysis

### Comparison with General Population

```{r comparison-analysis}
# YOUR CODE HERE: Compare children in care to all children
# This is a KEY analysis showing the gap
```

```{r plot-comparison}
# YOUR CODE HERE: Create comparison visualization
```

<!-- COMMENT: Discuss the gap and what it means -->

### Outcomes by Placement Type

```{r placement-analysis}
# YOUR CODE HERE: Analyze outcomes by placement type
# MUST USE PURRR HERE - this is a requirement!
# Example: Use map() or map_dbl() to analyze multiple groups
```

```{r plot-placement}
# YOUR CODE HERE: Visualize placement outcomes
```

<!-- COMMENT: Discuss which placements show better outcomes -->

### Temporal Trends (2018-2024)

```{r temporal-analysis}
# YOUR CODE HERE: Analyze trends over time
# Use purrr to analyze multiple years
```

```{r plot-trends}
# YOUR CODE HERE: Create trend visualization
```

<!-- COMMENT: Discuss how outcomes have changed over time -->

## Success Factors Analysis

### Predictive Analysis

```{r predictive-analysis}
# YOUR CODE HERE: Identify what predicts success
# This could use regression or similar approaches
# Use purrr for multiple model comparisons
```

```{r plot-predictors}
# YOUR CODE HERE: Visualize key predictors
```

<!-- COMMENT: Discuss which factors matter most -->

### Subgroup Comparisons

```{r subgroup-analysis}
# YOUR CODE HERE: Compare different subgroups
# Use purrr for efficient comparison across groups
```

<!-- COMMENT: Explain subgroup differences -->

## Summary Tables

```{r summary-tables}
# YOUR CODE HERE: Create comprehensive summary tables
# Include key statistics, rates, and comparisons
```

<!-- COMMENT: Explain what each table shows -->

## Part 1 Conclusions

<!-- 
YOUR CONTENT HERE:
- Summarize key findings from your analysis
- Answer your research questions
- Discuss policy implications
- Note any limitations
-->

\newpage

# Part 2: R Package Demonstration {#sec-part2}

<!-- 
YOUR CONTENT HERE:
- Choose an R package NOT used in the module
- Explain its purpose
- Demonstrate 3 main functions with ORIGINAL examples
- Cannot copy examples from help pages!
-->

## Package Overview

<!-- Introduce your chosen package -->

## Package Citation

```{r package-citation}
# YOUR CODE HERE: Get package citation
# citation("package_name")
```

## Function 1: [Function Name]

### Purpose

<!-- Explain what this function does -->

### Demonstration

```{r function1-demo}
# YOUR CODE HERE: Original example using the function
# Must be different from help page examples!
```

<!-- COMMENT: Explain the output and why it's useful -->

## Function 2: [Function Name]

### Purpose

<!-- Explain what this function does -->

### Demonstration

```{r function2-demo}
# YOUR CODE HERE: Original example
```

<!-- COMMENT: Explain the output -->

## Function 3: [Function Name]

### Purpose

<!-- Explain what this function does -->

### Demonstration

```{r function3-demo}
# YOUR CODE HERE: Original example
```

<!-- COMMENT: Explain the output -->

## Part 2 Summary

<!-- Summarize what the package is useful for -->

\newpage

# Part 3: Functions and Programming {#sec-part3}

<!-- 
YOUR CONTENT HERE:
- Create a statistical function
- Implement S3 classes
- Create print(), summary(), and plot() methods
- These methods must be VERY DIFFERENT from each other
- Include working examples
-->

## Function Overview

<!-- Explain what statistical analysis your function performs -->

## Main Function

```{r main-function}
# YOUR CODE HERE: Create your main analysis function
# It should return an object with a custom S3 class

# Example structure:
# my_analysis <- function(data, ...) {
#   # YOUR ANALYSIS LOGIC
#   
#   result <- list(
#     data = processed_data,
#     statistics = stats,
#     model = model_results,
#     # other components
#   )
#   
#   class(result) <- "your_class_name"
#   return(result)
# }
```

## Print Method

```{r print-method}
# YOUR CODE HERE: Create print method
# Should show simple, concise output

# print.your_class_name <- function(x, ...) {
#   cat("Your Analysis Results\n")
#   cat("=====================\n")
#   # YOUR PRINT LOGIC
# }
```

## Summary Method

```{r summary-method}
# YOUR CODE HERE: Create summary method
# Should show detailed statistical output
# Must be VERY DIFFERENT from print method!

# summary.your_class_name <- function(object, ...) {
#   # YOUR SUMMARY LOGIC
#   # Calculate additional statistics
#   # Return structured summary
# }
```

## Plot Method

```{r plot-method}
# YOUR CODE HERE: Create plot method
# Should create visualization of results

# plot.your_class_name <- function(x, ...) {
#   # YOUR PLOTTING LOGIC
#   # Create relevant visualization
# }
```

## Example Usage

```{r example-usage}
# YOUR CODE HERE: Demonstrate your function with real data

# Example:
# result <- my_analysis(data)
# 
# # Test print method
# print(result)
# 
# # Test summary method  
# summary(result)
# 
# # Test plot method
# plot(result)
```

## Part 3 Summary

<!-- Explain what your function does and how the methods work -->

\newpage

# Overall Conclusions

<!-- 
YOUR CONTENT HERE:
- Summarize findings from all three parts
- Discuss broader implications
- Limitations of the analysis
- Suggestions for future research
-->

\newpage

# References

<!-- References will auto-generate here from your .bib file -->

::: {#refs}
:::

\newpage

# Appendix: Additional Tables and Figures

```{r appendix-tables}
# YOUR CODE HERE: Additional supporting tables
```

```{r appendix-plots}
# YOUR CODE HERE: Additional supporting plots
```

---

<!-- 
SUBMISSION CHECKLIST:

Part 1 Analysis:
- [ ] Used at least one purrr function
- [ ] Only used packages from the module
- [ ] All code is commented
- [ ] Every plot and table is explained
- [ ] Clear introduction and conclusion
- [ ] Proper citations

Part 2 Package:
- [ ] Chose package not used in module
- [ ] Explained package purpose
- [ ] Demonstrated 3 functions
- [ ] Used original examples (not from help pages)
- [ ] Cited the package properly

Part 3 Functions:
- [ ] Created working function
- [ ] Assigned S3 class
- [ ] Created print() method
- [ ] Created summary() method (DIFFERENT from print!)
- [ ] Created plot() method
- [ ] Included working example

Quarto:
- [ ] Three sections clearly identifiable
- [ ] All code visible and commented
- [ ] No lines beyond margins
- [ ] File renders without errors
- [ ] Data file in same directory

Ready to submit:
- [ ] .Qmd file
- [ ] Rendered PDF
- [ ] Data files
-->
